Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-14 index.qmd
section:issues; process:write

<!-- 1. What does CJ do? -->
In *comparative judgment* (CJ) studies, judges assess a specific trait or competence across various stimuli by performing pairwise comparisons [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a]. Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to manifest a higher trait level. For example, when assessing text quality, judges compare pairs of written texts (the stimuli) to determine the relative quality each text exhibit (the trait) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 3. But what are the problems of DCJ as a method? -->
Nevertheless, despite the growing number of CJ studies, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed. This research primarily focuses on three: the apparent disconnect between CJ's measurement and structural model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role of comparison algorithms on the method's reliability and validity. The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aims to address all three concerns simultaneously.

@@

<!-- recording: Sven 24.10.04; time: 00:14:30 - 00:21:00 -->
<!-- recording: Sven and Tine 24.10.18; time: 00:08:30 - 00:14:55 -->
<!-- recording: Sven and Tine 24.10.25; time: 00:17:00 - 00:31:00 -->

<!-- 1. The BTL model is CJ's measurement model -->
In a typical CJ study, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] serves as the measurement model for CJ [@Andrich_1978; @Bramley_2008]. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

<!-- 3. But what is the problem with this? -->
However, the statistical literature advises caution when using estimated scores to conduct additional analyses or hypotheses tests. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty can introduce bias into the analysis and reduce the precision of hypothesis tests. Notably, the direction and magnitude of the bias are often unpredictable; results may be attenuated, exaggerated, or remain unaffected, depending on the amount of uncertainty present in the scores and the actual effects being tested [@Kline_et_al_2023, pp. 25; @Hoyle_et_al_2023, pp. 137]. Furthermore, reduced precision in hypothesis tests weakens their statistical power, ultimately increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].

<!-- A type-I error occurs when a *true* null hypothesis is incorrectly rejected, while a type-II error occurs when a *false* null hypothesis is incorrectly accepted [@Everitt_et_al_2010]. -->

<!-- 4. So, what is the solution?  -->
To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24] recommend conducting these analyses and tests within a structural model. A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010]. This approach allows analyses that can account for both the scores and their uncertainties simultaneously, rather than treating them as separate elements. Therefore, an integrated approach that combines CJ’s measurement and structural models can offer significant advantages.

## The assumptions of Case 5 and the measurement model

<!-- recording: Sven and Tine 24.10.25; time: 00:31:00 - 00:52:20 -->

<!-- 1. What are the five cases of Thurstone? -->

<!-- 1. the BTL model is the statistical articulation of Case 5 -->

<!-- From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927]. Talk about @Pollitt_et_al_2003 and @Bramley_2008. -->

<!--   What case 5 implies, Assumptions. Not a normal distribution but a logistic distribution -->

<!--   Table with assumptions -->

<!-- 2. But this model hold the greater number of assumptions -->

<!--   Although Case 5 was originally articulated to produce a "rather coarse scaling" of traits [@Thurstone_1927, p. 269], its assumptions have become predominant in the literature, but Thurstone's Case 5 is the case that holds the greater amount of simplifying assumptions. -->

<!--   @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are three: -->

<!--   (a) related to Case 2 and case 3, it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weighted 'mood' and 'simultaneous contrasts' effects. This is demonstrated using the additive nature of the logit scale, which helps to cancel 'bias' judges effects.  -->

@@

<!--   (b) related to Case 3, it assumes the correlation between stimuli is zero, this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli. Meaning there is an amount of correlation that it is not accounted for.  -->

@@

<!--   (c) related to case 5, discriminal dispersions of the stimuli are equal, but it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals. Example of exaggerated comparison of individuals, university students vs secondary school students. -->

@@

@@

<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015] -->

@@

>>>>

<!-- 1. What does CJ do? -->
In *comparative judgment* (CJ) studies, judges assess a specific trait or attribute across various stimuli by performing pairwise comparisons [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a]. Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher trait level. For example, when assessing text quality, judges compare pairs of written texts (the stimuli) to determine the relative quality each text exhibit (the trait) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 3. But what are the problems of DCJ as a method? -->
Nevertheless, despite the increasing number of CJ studies, unsystematic and fragmented research approaches have left several critical issues unaddressed. This research primarily focuses on three: the over-reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data, the apparent disconnect between CJ's trait measurement and hypothesis testing, and the unclear role and influence of comparison algorithms on the method's reliability and validity. The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aims to address all three concerns simultaneously.

@@

## The Case V and the statistical analysis of CJ data

<!-- recording: Sven and Tine 24.10.25; time: 00:31:10 - 00:52:20 -->
<!-- recording: Sven and Tine 24.11.12; time: 00:00:00 - 00:00:00 -->

<!-- 1. What is the general from of the CJ theory? -->
In its most general form, Thurstone's theory [-@Thurstone_1927] posits that when a judge compares two stimuli, the resulting dichotomous outcome depends on two main factors: the discriminal process and the law of comparative judgment. The *discriminal process* refers to the psychological effect each stimulus has on the judges, or more simply stated, the judges’ perception of the trait level of each stimulus. Thurstone assumes that the discriminal process for each stimulus follows a Normal distribution. In this distribution, the mode (mean), known as the *modal discriminal process*, represents the position of the stimulus on the trait continuum, while the dispersion, known as the *discriminal dispersion*, reflects variability in the perceived trait level of the stimulus. @fig-discriminal_process shows a visual depiction of the discriminal process of two stimuli (objects).

![Discriminal processes of two stimuli (objects). Extracted from @Bramley_2008 [pp. 249] ](/images/figures/discriminal_process.png){#fig-discriminal_process width=70%}

However, since the discriminal mode and dispersion of a single stimulus are not directly observable without comparing it to other stimuli, the *law of comparative judgment* becomes essential. This law asserts that when assessing a specific trait by comparing two stimuli, the stimulus positioned further along the continuum is perceived as having a higher level of that trait. Thus, the observed dichotomous outcome is determined by the distribution of the difference between the discriminal processes of the two stimuli, called the *discriminal difference*. @fig-comparative_judgment shows a visual depiction of the discriminal difference of two stimuli (objects).

![Discriminal difference between two stimuli (objects). Extracted from @Bramley_2008 [pp. 251] ](/images/figures/comparative_judgment.png){#fig-comparative_judgment width=65%}

<!-- 2. But several simplifying assumption are made to ease its use -->
Thurstone’s general form of the theory primarily applies to pairwise comparisons of stimuli made by a single judge [@Thurstone_1927, pp. 267]. Therefore, to enhance the theory's practical applicability, he developed five distinct cases, each incorporating a series of increasingly simplifying assumptions. @fig-thurstone_cases outlines each case along with their key assumptions. For a detailed explanation of this theoretical progression, see @Thurstone_1927 and @Bramley_2008 [pp. 248-253].

![Thurstones cases and asumptions](/images/tables/thurstone_cases.png){#fig-thurstone_cases width=85%}

<!-- 3. And yet Case V is is the most commonly used -->
Surprisingly, Case V is the most commonly used case in the CJ literature, despite incorporating the largest number of simplifying assumptions [@Bramley_2008, pp. 253]. This widespread adoption is primarily due to the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959], which offers a simpler statistical representation of the case. The BTL model closely aligns with Case V’s assumptions, with one key difference: the distribution of the discriminal process. While Case V assumes a Normal distribution, the BTL model uses the more mathematically tractable logistic distribution [@Bramley_2008, pp. 254] (see @fig-thurstone_cases). However, this adjustment does not significantly affect the model’s estimation or interpretation, as the logistic and normal distributions are related to each other by a factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16].

<!-- the mathematical function linking dichotomous outcomes to trait scores -->

<!-- 4. But two assumptions of Case V might not hold -->
However, because Case V was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927, pp. 269], prioritizing statistical ease of use over precision in trait measurement, two of its assumptions may be causing more problems than expected: the assumption of zero correlation between the stimuli, and the assumption of equal dispersion between stimuli.

<!-- @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are two: -->

<!--   (a) it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weighted 'mood' and 'simultaneous contrasts' effects. This is demonstrated using the additive nature of the logit scale, which helps to cancel 'bias' judges effects.  -->

@@

<!--   (b) it assumes the correlation between stimuli is zero, this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli. Meaning there is an amount of correlation that it is not accounted for.  -->

@@

<!-- (c) discriminal dispersions of the stimuli are equal, but it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals. Example of exaggerated comparison of individuals, university students vs secondary school students. -->

@@

<!-- "It is probably true that this variability of the discriminal dispersions on the psychological continuum is of relatively less serious importance in dealing with strictly homogenous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogenous stimulus series such as handwriting specimens, English compositions, sewing samples, oriental rugs." [@Thurstone_1927]. -->

@@

<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp. 2] -->

<!-- recording: Sven 24.10.04; time: 00:14:30 - 00:21:00 -->
<!-- recording: Sven and Tine 24.10.18; time: 00:08:30 - 00:14:55 -->
<!-- recording: Sven and Tine 24.10.25; time: 00:17:00 - 00:31:10 -->

<!-- 1. The BTL model is CJ's measurement model -->
As outlined in the previous section, the BTL model typically functions as the measurement model in a CJ study [@Andrich_1978; @Bramley_2008]. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

<!-- 3. But what is the problem with this? -->
However, the statistical literature advises caution when using estimated scores to conduct additional analyses or hypotheses tests. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty can introduce bias into the analysis and reduce the precision of hypothesis tests. Notably, the direction and magnitude of the bias are often unpredictable; results may be attenuated, exaggerated, or remain unaffected, depending on the amount of uncertainty present in the scores and the actual effects being tested [@Kline_et_al_2023, pp. 25; @Hoyle_et_al_2023, pp. 137]. Furthermore, reduced precision in hypothesis tests weakens their statistical power, ultimately increasing the likelihood of committing type-I or type-II errors [@McElreath_2020]. 
<!-- A type-I error occurs when a *true* null hypothesis is incorrectly rejected, while a type-II error occurs when a *false* null hypothesis is incorrectly accepted [@Everitt_et_al_2010]. -->

<!-- 4. So, what is the solution?  -->
To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24] recommend conducting these analyses and tests within a structural model. A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010]. This approach allows analyses that can account for both the BTL scores and their uncertainties simultaneously, rather than treating them as separate elements. Therefore, an integrated approach that combines CJ’s measurement and structural models can offer significant advantages.

@@

<!-- It is assumed that the single observer compares each pair of stimuli a sufficient number of times so that a proportion may be determined for eah pair [@Thurstone_1927]. -->
